---
title: "Association analysis of compositional microbiome data using R packages"
author: "David Xin Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) {
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_file=file.path(dirname(inputFile), out_dir, 'index.html'))})
output:
  html_document:
    # theme: cosmo
    highlight: pygments
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    collapsed: FALSE
    number_sections: FALSE
    fig_width: 7
    fig_height: 6
    fig_caption: TRUE
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

<html>

<head>

```{=html}
<style>

h1{
 color: #055C9D;
 font-family: Georgia;
 font-size: 200%
}


h2{
 color: #055C9D;
 font-family: helvetica;
 font-size: 150%
}

h3{
 color: #055C9D;  
 font-family: helvetica;
 font-size: 120%; 
}

p {
 color: #333333;
 font-family: helvetica;
 font-size: 100%;
}

</style>
```
</head>

</html>

```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.dim =c(6,4), 
                      dev="png",
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      message=FALSE) #stop render when error occurs   
```

## Introduction

Association analysis in microbiome studies infers relationship between
OTUs/ taxa. Pearson r Correlation and Spearman's Rank Correlation are
classical methods in association analysis. However, microbiome relative
data is compositional and consequently make the classical methods
inappropriate. If naively applying classical methods to compositional
microbiome data, we likely end up with getting spurious correlations.

The review article published in 2017, "Microbiome datasets are
compositional: and this is not optional" [@gloor2017]summarized four
newly-invented statistical methods dealing with compositionality and
achieving reliable association relationship. The four methods consists
of SparCC, Spiec Easi, proportionality (ϕ and ρ).

Proportionality is a precise indicator of absolute correlation, although
sensitivity is limited [@quinn2017].

In this project, I intend to demonstrate analysis over the HMP-IBD
microbiome dataset using the above four statistical methods; and to
compare outcomes based on the four methods against ones based on
Spearman's Rank Correlation.

## R workflow (use proportionality analysis with propr package as an example)

1.  Load dataset
2.  Calculate proportionality
3.  Identify proportionally abundant taxa
4.  Visualizing proportionality
5.  Differential proportionality analysis

## Scripts

```{r libaries}
library(tidyverse) 

library(devtools)
install_github("tpq/propr")  
library(propr)

library(stringr)

```

### Step 1. Load dataset

The microbiome OTU table and metadata was retrieved from [ML
Repo](https://knights-lab.github.io/MLRepo/docs/turnbaugh_lean_obese_all.html)

```{r}

# raw OTU table
raw_otu <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/refseq/otutable.txt",
                    header=T,
                    sep = "")

```

The raw OTU table contains 557 features (OTUs) in 281 samples.

Shorten OTU identifiers with keeping only genus and species names, using
`stringr` package. And then accumulate counts of OTUs at the species
level.

```{r shorten OTU names}

head(raw_otu$X.OTU) 

# split strings 
otu.id <- raw_otu$X.OTU

split_otu.id <- str_split_fixed(otu.id, "_", n=6) %>% 
        as.data.frame()

# join genus and species fragments 
join_split_otu.id <- split_otu.id %>% 
        mutate(taxa = str_c(V3, V4, sep = " ")) 

taxa <- join_split_otu.id$taxa 

# merge taxa to raw otu table 
raw_otu_taxa <- cbind(raw_otu, taxa) 


# accumulate counts at the species level 
otu_count_splevel <- raw_otu_taxa %>% 
        gather(key = "sample", value = "count", -c("ID", "X.OTU", "taxa")) %>% 
        group_by(sample, taxa) %>%
        summarise(count_sp = sum(count, na.rm = T)) %>% 
        arrange(desc(count_sp)) %>%
        ungroup()

```

Read in the metadata from URL.

```{r}

# metadata 

meta <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/task-obese-lean-all.txt",
         header = TRUE,
         sep = "") %>% 
        mutate(sample = X.SampleID, is.obese = Var) %>% 
        select(sample, is.obese)

head(meta)

```

The metadata contains `r dim(meta)[1]` samples. `Var` indicates the
independent, binary variable of interest, `Lean` and `Obese`.


Transpose the OTU table into sample x taxa format which is suitable for
proportionality analysis.

```{r transpose OTU table}
otu_count_splevel2 <- otu_count_splevel %>% 
        spread(key = "taxa", value = count_sp) %>%
        column_to_rownames("sample")

```

Now, OTU count table at the species level contains 517 features, in 281
samples.

Filter low abundant taxa before the proportionality analysis. Keep only
the taxa with at least 10 counts in at least 5 samples.

```{r filter low abundant taxa}

keep_index <- apply(otu_count_splevel2, 2, function(x) sum(x >= 10) >= 5)  


```
The resulting OTU table after filtering the low abundant taxa contains 154 unique species/taxa. 


Add group variables to the resulting OTU table. 
```{r group variables}

otu_count_splevel2_meta <- otu_count_splevel2 %>%
        rownames_to_column("sample") %>%
        inner_join(meta, by= "sample") %>%
        column_to_rownames("sample") 

head(otu_count_splevel2_meta)

```


Check if any missing values in the OTU table. Replace missing values
with zero if any.

```{r handle NA}

sum(is.na(otu_count_splevel2)) 

# otu_count_splevel2_rmna <- otu_count_splevel2 %>%
#         replace(is.na(.), 0)
# 
# sum(is.na(otu_count_splevel2_rmna))

```

### Step 2. Calculate proportionality using propr R package

Calculate a proportionality measurement, `rho` using `propr` function,
following the procedures and codes provided by T.P.Quinn et al.
[@quinn2019]

Set three required arguments in the wrapper `propr` function as follows,

-   `counts` a count matrix with subjects as rows and features as
    columns

-   `metric` the proportionality metric to calculate, 'rho', 'pho' or
    'phs'

-   `ivar` reference OTU/ feature

```{r proportionality}

pr <- propr(counts = otu_count_splevel2, # a count matrix with subjects as rows and features as columns
            metric = "rho", # the proportionality metric to caculate, 'rho', 'pho' or 'phs'
            ivar = "clr", # reference OTU/ feature for alr
            symmetrize = FALSE,
            p = 100, # permutation cycles (100, by default)  
            select = keep_index) # filter low-abundant taxa 


## alternatively, calculate phi, rho or phs using phit(), perb() and phis(), respectively. 
# phi <- phit(raw_otu_t_na, symmetrize = TRUE) 
# 
# rho <- perb(raw_otu_t_na, ivar = 0)
# 
# phs <- phis(raw_otu_t_na, ivar = 0) 

```

For proportionality, permute false discovery rate (FDR) for a given
cut-off, instead of estimating parametric P-value.

```{r permutate FDR}

# select a good cutoff for 'rho' by permuting the FDR at various cutoffs. 
# below, use [0, .05, ..., .95, 1]

# pr <- updateCutoffs(pr, cutoff = seq(0,1,.05))   
# 
# pr@fdr

```

According to the package
[vignette](https://rdrr.io/cran/propr/f/vignettes/a_introduction.Rmd),
choose the largest cutoff that keeps the FDR below 0.05. In this case,
`rho` cutoff is 0.20.

Visualize proportionality with a strict cutoff, using `getNetwork`
function. The package
[vignette](https://rdrr.io/cran/propr/f/vignettes/b_visualization.Rmd)
describes several built-in tools for visualizing proportionality.

```{r proportionality-based network}

getNetwork(pr, cutoff = 0.2) 


```

Export highly-proportional (rho \>= 0.9) features (OTU).

```{r highly proportional pairs}

result_rho_df <- getResults(pr, cutoff = 0.2) 

result_rho_df 

```

How to interpret proportionality? Proportionality depends on
log-transformation and must get interpreted with respect to the chosen
reference.

Interpret clr-based proportionality to signify a coordination that
follows the general trend of the data. In other words, these
proportional OTUs move together as individuals relative to how most
genes move on average [@quinn2019].

## Differential proportionality analysis

`propd` functions in the `propr` package can test whether
proportionality of certain pairs change between binary experimental
groups.

According to the package
[vignette](https://cran.microsoft.com/snapshot/2018-04-24/web/packages/propr/vignettes/e_differential.html),
`propd` method compares the variance of log-ratio (VLR) for one pair
across groups, considering two types of differential proportionality,

-   `disjointed proportionality`: proportionality of a pair holds in
    both groups, but the ratio between the partners changes between the
    groups (i.e., the slope of the proportionality changes)

-   `emergent proportionality`: proportionality exits in only one of the
    groups (i.e., the strength of the proportionality changes)
    
`propd` function estimates differential proportionality by calculating $theta$ for all features pairs. 

The function takes the following arguments as input:

- `counts` a matrix of $n$ samples (as rows) and $d$ features (as columns) 

- `group` an $n$-dimensional vector corresponding to subject labels 

- `alpha` an optional argument to trigger and guide transformation 

- `p` the total number of permutations used to estimate FDR 

```{r differential proportionality}

counts <- otu_count_splevel2_meta %>% select(-is.obese)

group <- otu_count_splevel2_meta$is.obese 

pd <- propd(counts, group, alpha = NA, p=100) 

```

```{r thetad}
 
theta_d <- setDisjointed(pd)  # disjointed diff prop 

theta_e <- setDisjointed(pd)  # emergent diff prop 

```


```{r tabulate results}

theta_d_df <- getResults(theta_d) 

theta_d_df %>% 
  arrange(theta) 


theta_e_df <- getResults(theta_e) 

theta_e_df %>% 
  arrange(theta_e)

```


```{r plot}

plot(theta_d@counts[, "Odoribacter splanchnicus"], 
     theta_d@counts[, "Bacteroides massiliensis"], 
     col = ifelse(theta_d@group == "Obese", "red", "blue"))
grp1 <- theta_d@group == "Obese"
grp2 <- theta_d@group != "Obese"
abline(a = 0, 
       b = theta_d@counts[grp1, "Odoribacter splanchnicus"] / theta_d@counts[grp1, "Bacteroides massiliensis"], 
       col = "red")
abline(a = 0, 
       b = theta_d@counts[grp2, "Odoribacter splanchnicus"] / theta_d@counts[grp2, "Bacteroides massiliensis"], 
       col = "blue")


```

```{r plot ratio abundance}


plot(theta_d@counts[, "Odoribacter splanchnicus"] / theta_d@counts[, "Bacteroides massiliensis"],
     col = ifelse(theta_d@group == "Obese", "red", "blue"))


```
visualize network based on differential proportionality
```{r }

g <- plot(theta_d, 
          cutoff=300, 
          d3= F)  # TRUE for displaying 3-D plot 

```



```{r emergent diff network}

plot(theta_e, cutoff=200, d3=T) 



```
