---
title: "Proportionality analysis of compositional microbiome data, using a real dataset"
author: "Xin (David) Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) {
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_file=file.path(dirname(inputFile), out_dir, 'index.html'))})
output:
  html_document:
    # theme: cosmo
    highlight: pygments
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    collapsed: FALSE
    number_sections: TRUE
    fig_width: 7
    fig_height: 6
    fig_caption: TRUE
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

<html>

<head>

```{=html}
<style>

h1{
 color: #055C9D;
 font-family: Georgia;
 font-size: 200%
}


h2{
 color: #055C9D;
 font-family: helvetica;
 font-size: 150%
}

h3{
 color: #055C9D;  
 font-family: helvetica;
 font-size: 120%; 
}

p {
 color: #333333;
 font-family: helvetica;
 font-size: 100%;
}

.blackbox {
  padding: 1em;
  background: green;
  color: black;
  border: 2px solid orange;
  border-radius: 10px;
}

.center {
  text-align: center;
}

</style>
```

</head>

</html>

```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.height=6, fig.width=8,
                      dev="png",
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      collapse = FALSE, 
                      message=FALSE) #stop render when error occurs   
```

## Project purpose

This mini project aims to:

-   demonstrate the proportionality analysis on a real human dataset,
    using `propr` R package

-   validate the proportionality method by evaluating precision and
    recall

## Problem

Association analysis in microbiome studies infers relationship between
OTUs/ taxa. Pearson r Correlation and Spearman's Rank Correlation are
classical methods in association analysis. However, microbiome relative
data is compositional and consequently make the classical methods
inappropriate. If naively applying classical methods to compositional
microbiome data, we will get spurious correlations.

## Solution

The review article published in 2017, "Microbiome datasets are
compositional: and this is not optional" [@gloor2017]summarized four
valid statistical methods dealing with compositionality and achieving
reliable association relationship. The four methods consists of SparCC,
Spiec Easi, proportionality (ϕ and ρ).

Proportionality is a precise indicator of absolute correlation, although
sensitivity is limited [@quinn2017].

## Markdown outline

-   Calculate proportionality
-   Identify highly proportional taxa-pairs
-   Visualize proportionality
-   Evaluate precision and recall
-   Analyze differential proportionality

## Scripts

```{r libaries, results='hide', warning=FALSE}

library(tidyverse) 

library(devtools)
# install_github("tpq/propr")  
library(propr)
library(Hmisc)  # Spearman correlation 
library(corrplot)  # visualize correlation matrix 
library(stringr)

```

### Load datasets

The microbiome OTU table and metadata was retrieved from [ML
Repo](https://knights-lab.github.io/MLRepo/docs/turnbaugh_lean_obese_all.html)

```{r}

# raw OTU table
raw_otu <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/refseq/otutable.txt",
                    header=T,
                    sep = "")

```

The raw OTU table contains 557 features (OTUs) in 281 samples.

Shorten OTU identifiers with keeping only genus and species names, using
`stringr` package. And then accumulate counts of OTUs at the species
level.

```{r shorten OTU names}

head(raw_otu$X.OTU) 

# split strings 
otu.id <- raw_otu$X.OTU

split_otu.id <- str_split_fixed(otu.id, "_", n=6) %>% 
        as.data.frame()

# join genus and species fragments 
join_split_otu.id <- split_otu.id %>% 
        mutate(taxa = str_c(V3, V4, sep = " ")) 

taxa <- join_split_otu.id$taxa 

# merge taxa to raw otu table 
raw_otu_taxa <- cbind(raw_otu, taxa) 


# accumulate counts at the species level 
otu_count_splevel <- raw_otu_taxa %>% 
        gather(key = "sample", value = "count", -c("ID", "X.OTU", "taxa")) %>% 
        group_by(sample, taxa) %>%
        summarise(count_sp = sum(count, na.rm = T)) %>% 
        arrange(desc(count_sp)) %>%
        ungroup()

```

Read in the metadata from URL.

```{r}

# metadata 

meta <- read.csv(file = "https://knights-lab.github.io/MLRepo/datasets/turnbaugh/task-obese-lean-all.txt",
         header = TRUE,
         sep = "") %>% 
        mutate(sample = X.SampleID, is.obese = Var) %>% 
        select(sample, is.obese)

head(meta)

```

The metadata contains `r dim(meta)[1]` samples. `Var` indicates the
independent, binary variable of interest, `Lean` and `Obese`.

### 2. Preprocess data frames for following proportionality analysis

Transpose the OTU table into sample x taxa format which is suitable for
proportionality analysis.

```{r transpose OTU table}
otu_count_splevel2 <- otu_count_splevel %>% 
        spread(key = "taxa", value = count_sp) %>%
        column_to_rownames("sample")

```

Now, the resulting OTU count table at the species level contains 517
features/ species, in 281 samples.

Identify low-abundant OTUs and index them in the OTU table.

```{r filter low abundant taxa}

keep_index <- apply(otu_count_splevel2, 2, function(x) sum(x >= 10) >= 5)  # index of low abundant OTUs 

```

154 unique species/taxa remains after dropping the low-abundant taxa.

Add group variables to the resulting OTU table.

```{r group variables}

otu_count_splevel2_meta <- otu_count_splevel2 %>%
        rownames_to_column("sample") %>%
        inner_join(meta, by= "sample") %>%
        column_to_rownames("sample") 

head(otu_count_splevel2_meta) # print first rows 

# knitr::kable(
#   head(otu_count_splevel2_meta[ ,1:8], 10),
#   booktabs=TRUE,
#   caption = "A table of the first 6 rows of the OTU matrix at the species level"
# )

```

Check if any missing values in the OTU table. Replace missing values
with zero if any.

```{r handle NA}

sum(is.na(otu_count_splevel2)) 

# otu_count_splevel2_rmna <- otu_count_splevel2 %>%
#         replace(is.na(.), 0)
# 
# sum(is.na(otu_count_splevel2_rmna))

```

### Calculate proportionality

Calculate a proportionality measurement, `rho` using `propr` function,
following the procedures and codes provided by T.P.Quinn et al.
[@quinn2019]

```{r proportionality}

# calculate the propr object 
pr <- propr(counts = otu_count_splevel2, # a count matrix with subjects as rows and features as columns
            metric = "rho", # the proportionality metric to calculate, 'rho', 'pho' or 'phs'
            ivar = "clr", # reference OTU/ feature for alr. 
            symmetrize = FALSE,
            p = 100) # permutation cycles (100, by default)  
            # select = keep_index) # filter low-abundant taxa

# pr@matrix # proportionality metrics 

# subset the propr object, dropping low-abundant taxa 
pr_sub <- subset(pr, select = keep_index)

# pr_sub@matrix # proportionality metrics of high-enough-abundant taxa 


# the alternative way to calculate rho statistics
# rho <- perb(otu_count_splevel2,
#             ivar =  "clr",
#             select= keep_index)  # subset abundant taxa after calculating proportionality on the complete matrix


```

::: {.blackbox data-latex=""}
::: {.center data-latex=""}
Set three required arguments in the wrapper `propr` function as follows,
:::

-   `counts` a count matrix with subjects as rows and features as
    columns

-   `metric` the proportionality metric to calculate, 'rho', 'pho' or
    'phs'

-   `ivar` reference OTU/ feature
:::

For proportionality, permute false discovery rate (FDR) for a given
cut-off, instead of estimating parametric P-value.

```{r permutate FDR}

# select a good cutoff for 'rho' by permuting the FDR at various cutoffs.
# below, use [0, .05, ..., .95, 1]

pr <- updateCutoffs(pr, cutoff = seq(0, 1,.05)) 

# pr@fdr  # identify the maximal cutoff at FDR < 5% 


```

The table shows that the largest cutoff is 0.20 when the FDR reaches 5%
or fewer.

Identify highly-proportional (rho \>= 0.7) feature/ species pairs.

```{r highly-proportional-pairs}

pr_sub_high <- pr_sub[">", .7] 
pr_sub_high 

pr_sub_high@pairs # index of highly proportional pairs 

pr_sub_high_mt <- pr_sub_high@matrix %>% 
  as.data.frame()

high_propr_pairs <- pr_sub_high_mt %>% 
  rownames_to_column("partner") %>%
  mutate(partner = factor(partner)) %>% 
  gather(key="pair", value = "rho", -partner) %>% 
  mutate(pair = factor(pair)) %>% 
  mutate(partner_index = as.numeric(partner)) %>% 
  mutate(pair_index = as.numeric(pair)) %>% 
  filter(partner_index < pair_index) %>% 
  filter(rho > 0.7) %>% 
  select(partner, pair, rho) %>% 
  mutate(rho = round(rho, 2))


knitr::kable(high_propr_pairs, 
             booktab = TRUE,
             caption = "Highly proportional species pairs (rho > 0.7)") 
  
```

### Visualize proportionality

Visualize proportionality with a strict cutoff, using `getNetwork`
function. The package
[vignette](https://rdrr.io/cran/propr/f/vignettes/b_visualization.Rmd)
describes several built-in tools for visualizing proportionality.

```{r proportionality-based network, message=FALSE, fig.height=6, fig.width=8, fig.cap="Proportionality network based on rho matrics"}

getNetwork(pr, 
           cutoff = 0.7, 
           d3=F) # not display in 3D in this case 

# plot(pr_sub_high)
```

### Interprete proportionality

How to interpret proportionality? Proportionality depends on
log-transformation and must get interpreted with respect to the chosen
reference.

Interpret clr-based proportionality to signify a coordination that
follows the general trend of the data. In other words, these
proportional OTUs move together as individuals relative to how most
genes move on average [@quinn2019].

### Evaluate precision and recall (sensitivity) of the proportionality method

Apply Spearman rank correlation to raw absolute abundance matrix.
Consider the resulting Spearman correlation coefficients (absolute
correlation) as the true correlations between taxa pairs.

Use `Hmisc` package to implement Spearman rank correlation analysis and
`corrplot` package to visualize correlation matrix.

```{r spearman-absoulte-corr, fig.height=6, fig.width=8, message=FALSE, fig.cap="Correlation matrix of 517 species based on Spearman rank correlation method"}

spearman_corr <- Hmisc::rcorr(as.matrix(otu_count_splevel2), 
                              type = "spearman")   

corrplot::corrplot(spearman_corr$r, 
                   method = "color", 
                   tl.pos = "n",  # mutate text labels 
                   type="upper") 

```

```{r compare-abs-vs-rel-corr}

# extract Spearman coefficients 
abs_corr_r <- spearman_corr$r %>% 
  as.data.frame() %>% 
  rownames_to_column("partner") %>% 
  mutate(partner = factor(partner)) %>% 
  gather(key="pair", value = "sp_coef", -partner) %>% 
  mutate(pair = factor(pair)) %>% 
  mutate(partner_index = as.numeric(partner)) %>% 
  mutate(pair_index = as.numeric(pair)) %>% 
  filter(partner_index < pair_index) %>%
  arrange(partner_index, pair_index)

# extract Spearman p-val 
abs_corr_pval <- spearman_corr$P %>% 
  as.data.frame() %>% 
  rownames_to_column("partner") %>% 
  mutate(partner = factor(partner)) %>% 
  gather(key="pair", value = "sp_pval", -partner) %>% 
  mutate(pair = factor(pair)) %>% 
  mutate(partner_index = as.numeric(partner)) %>% 
  mutate(pair_index = as.numeric(pair)) %>% 
  filter(partner_index < pair_index) %>% 
  arrange(partner_index, pair_index) 
  
# combine Spearman coefficients and p-val 
abs_corr_r_pval <- abs_corr_r 

abs_corr_r_pval$sp_pval <- abs_corr_pval$sp_pval 

# proportionality matrix (rho)
propr_rho <- pr@matrix %>%
  as.data.frame() %>% 
  rownames_to_column("partner") %>%
  mutate(partner = factor(partner)) %>% 
  gather(key="pair", value = "rho", -partner) %>% 
  mutate(pair = factor(pair)) %>% 
  mutate(partner_index = as.numeric(partner)) %>% 
  mutate(pair_index = as.numeric(pair)) %>% 
  filter(partner_index < pair_index) %>% 
  arrange(partner_index, pair_index)

# combine absolute correlation and relative correlation coefficients 

abs_rel_coef <- abs_corr_r_pval 

abs_rel_coef$rho_propr <- propr_rho$rho 

ggplot(abs_rel_coef, aes(x= sp_coef, y= rho_propr))+
  geom_jitter(alpha = 0.5) +
  stat_smooth(method = "lm")


# contingency table 
cont_matrix <- table("Actual" = abs_rel_coef$sp_coef > 0.9,
                     "Observed" = abs_rel_coef$rho_propr > 0.9)  

cont_matrix # print 


prop.table(cont_matrix, 1)[1,1] # specificity
prop.table(cont_matrix, 1)[2,2] # sensitivity 


```

At the rho (proportionality matrices) cutoff of 0.9, the specificity of
the proportionality is 91.1% and the sensitivity 62.5%.

### Differential proportionality analysis

`propd` functions in the `propr` package can test whether
proportionality of certain pairs change between binary experimental
groups.

The package
[vignette](https://cran.microsoft.com/snapshot/2018-04-24/web/packages/propr/vignettes/e_differential.html)
explains mathematically the differential proportionality measures.

According to the package
[vignette](https://cran.microsoft.com/snapshot/2018-04-24/web/packages/propr/vignettes/e_differential.html),
`propd` method compares the variance of log-ratio (VLR) for one pair
across groups, considering two types of differential proportionality,

-   `disjointed proportionality`: proportionality of a pair holds in
    both groups, but the ratio between the partners changes between the
    groups (i.e., the slope of the proportionality changes)

-   `emergent proportionality`: proportionality exits in only one of the
    groups (i.e., the strength of the proportionality changes)

`propd` function estimates differential proportionality by calculating
$theta$ for all features pairs.

The function takes the following arguments as input:

-   `counts` a matrix of $n$ samples (as rows) and $d$ features (as
    columns)

-   `group` an $n$-dimensional vector corresponding to subject labels

-   `alpha` an optional argument to trigger and guide transformation

-   `p` the total number of permutations used to estimate FDR

```{r differential-proportionality-estimation}

counts <- otu_count_splevel2_meta %>% select(-is.obese) # count matrix, sample X taxa 

group <- otu_count_splevel2_meta$is.obese # binary classes, obese vs lean 

pd <- propd(counts, group, alpha = NA, p=100) # calculate disjointed- and emergent propr. 

```

According the package vignette, the propd object contains theta_d and
theta_e (among others), although only theta_d is active by default.
Users can easily change which theta is active the functions
`setDisjointed` and `setEmergent`.

```{r disjointed-and-emergent-differential-propr, collapse=TRUE}
 
theta_d <- setDisjointed(pd)  # activate only disjointed diff prop 

theta_e <- setEmergent(pd)  # activate only emergent diff prop 

```

Extract the detailed results from the above objects, including OTU pairs
and differential proportionality statistics.

How to identify most important pairs based on the theta_d and theta_e?
Qualitatively, the smaller the theta_d, the more important the
proportional pairs are. In contrast, the bigger the theta_e, the more
important the proportional pairs are.

```{r estimate-FDR-disjointed-propr}

# estimate FDR using updateCutoffs function
theta_d <- updateCutoffs(theta_d, cutoff = seq(0.91, 1, 0.001)) 

theta_d@fdr  

knitr::kable(head(theta_d@fdr, 10),
             caption = "FDRs at various rho cutoffs",
             digits = 2) 
```

```{r disjointed-diff-propr-output}

# disjointed diff. proportionality 
theta_d_df <- getResults(theta_d) 

```

The column `theta` equals theta_d matrices. Values of `theta` are 0.91
or greater which are relatively large, indicating that almost all
proportional pairs do not obviously change with experimental conditions
(i.e., obese or not). In addition, FDR of all the proportional pairs are
greater than 5%.

Similarly, estimate FDR for the emergent differential proportionality.

```{r estimate-FDR-emergent-diff, results='hide'}

theta_e <- updateCutoffs(theta_e)

```

```{r emergent-diff-propr-output}

# emergent diff. proportionality 
theta_e_df <- getResults(theta_e)  

```

```{r disjointed-prop-based-scatter-plot, fig.height=6, fig.width=8, include=FALSE}

plot(theta_d@counts[, "Odoribacter splanchnicus"],
     theta_d@counts[, "Bacteroides massiliensis"],
     col = ifelse(theta_d@group == "Obese", "red", "blue"))
grp1 <- theta_d@group == "Obese"
grp2 <- theta_d@group != "Obese"
abline(a = 0,
       b = theta_d@counts[grp1, "Odoribacter splanchnicus"] / theta_d@counts[grp1, "Bacteroides massiliensis"],
       col = "red")
abline(a = 0,
       b = theta_d@counts[grp2, "Odoribacter splanchnicus"] / theta_d@counts[grp2, "Bacteroides massiliensis"],
       col = "blue")


```

```{r scatter-plot-ratio-abundance, fig.height=6, fig.width=8, include=FALSE}

plot(theta_d@counts[, "Odoribacter splanchnicus"] / theta_d@counts[, "Bacteroides massiliensis"],
     col = ifelse(theta_d@group == "Obese", "red", "blue"))

```

### Disproportionality networks

Visualize the network based on disjointed differential proportionality
using `plot` function.

Providing a value of [0,1] to the cutoff argument will index pairs based
on a maximum value of theta. Alternately, providing an integer greater
than 1 to the cutoff argument will index the top N pairs as ranked by
theta.

Note that setting `d3 = TRUE` will have the `rgl` package render the
network in 3D.

```{r disjointed-prop-network, fig.height=6, fig.width=8, fig.cap="Disproportionality network based on disjointed proportionality"}

net_d <- plot(theta_d, # disjointed proportionality object 
              cutoff= 300,  
              d3= F)  # whether or not display 3-D plot 

```

Each "node" represents a species/taxa while each connecting "edge"
represents an indexed pair (i.e. theta \< cutoff).

For disjointed proportionality networks, red edges show an increase in
log-ratio means in Obese compared to Lean while blue edges show a
decrease in log-ratio means in Lean.

Importantly, a small number of taxa/ species participants in a large
number of the top differentially proportional pairs.

```{r taxa-in-disjointed-propr, include=FALSE}

net_d

```

Visualize the network based on emergent differential proportionality
using `plot` function.

```{r emergent-prop-network, fig.height=6, fig.width=8, fig.cap="Disproportionality network based on emergent proportionality" }

net_e <- plot(theta_e, # emergent proportionality object 
              cutoff=800, 
              d3=F)  # whether or not display 3-D plot   

```

## Package version 
The analysis was performed with R version `r getRversion()`. Version information of all the R packages are as follow, 
```{r package version}

sessionInfo()

```



